
```{r ch1setup, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(GGally)
options(dplyr.summarise.inform = FALSE,
        warn = -1)
```


# Introductory Statistics in R

We begin our dive into Statistical Modeling by first reviewing content from introductory statistics and introduce fundamental elements of the R programming language [@R-base]. The learning objective of this chapter include:

* Review of the fundamental concepts of *populations*, *samples*, *parameters* and *statistics*.
* Review of exploratory and descriptive statistics: numerical quantities and graphs.
* An introduction and overview of the R software for statistical computing.
* Basic statistical inference through a two-sample $t$-test and confidence intervals.

## Goals of a statistical analysis

The field of **Statistics** can broadly be defined as the study of variation. 

We collect **sample** data from a parent **population** of interest. Ultimately, we are interested in characteristics or **parameters** about the **population**. We use a **sample** because it is unfeasible or overly costly to collect data from every member of a **population** (that process is known as a *census*). 

Good statistical practice demands that we collect the  **sample** data from the **population** in such a manner so as to ensure that the **sample** is representative; *i.e.*, it is not presenting a systematic bias in its representation. With a *good* sample, we can make decisions about the population using that sample.

There are two main sources of variation that arise in statistical analysis:

1. Measurements of the same attribute will vary from observation to observation. This is natural and is to be expected. Consider how your height may vary compared to your parents and siblings. This *random* (or *stochastic*) component is typically modeled through probabilistic assumptions (think the *Normal* distribution from your Intro Stat course). This component is often called *noise* or *error*, although the later is a misnomer. 

2. We are interested in some **parameter** that helps describes characteristics of the population. Any characteristic computed from the sample is known as a sample **statistic**. Ideally, this **statistic** will reasonably approximate the **parameter** of interest. Since only a subset of the population is used in the calculation of a **statistic**, it will *vary* from their corresponding population **parameter**. In fact, different samples from the same population will produce different estimates of the same parameter, so sample statistics also vary from sample to sample.

This variation gives rise to the concept of uncertainty in any findings we make about a population based on sample data. A meaningful statistical analysis helps quantify this uncertainty.

The **goals** of a statistical analysis typically are

* to make sense of data in the face of uncertainty,
* to meaningfully describe the patterns of variation in collected sample data, 
   + and use this information to make reliable inferences/generalizations about the parent population.

### Conducting a statistical analysis

The practice of statistics starts with an underlying problem or contextual question. Prior to performing an analysis, we must that original question. Some examples may include:

* Are fish sensitive to Ultraviolet light?
* What socio-economic variables and demographics predict college graduation rates?
* Can we determine if there truly is a *home field* advantage in sports?
* Does a new experimental drug improve over an existing drug?

From there, data is collected through an experiment, by observation or via a survey. You then proceed with a data analysis, and then finish with statistical and contextual conclusions. 

Unfortunately, it is common for inexperienced statisticians to start with a complex analysis without ever considering the objectives of the study. It is also common for inexperienced statisticians to not consider whether the data are appropriate for the chosen analysis and whether the underlying question can be answered.

We recommend you follow the guidance (outlined below) provided in Chapter 1 of the text by @Faraway2004linear -- Look before you leap!

**Problem formulation** - To perform statistical analysis correctly, you should:

1. **Understand the context**. Statisticians often work in collaboration with researchers in other disciplines and need to understand something about the subject area. You should regard this as an opportunity to learn something new. In essence, a Statistician must be a renaissance person -- some level of *elementary* expertise is needed in the subject area.

2. **Understand the objectives**. Again, often you will be working with a collaborator who may not be clear about their objectives. Ideally they approach a statistician before data collection. This helps alleviate several potential issues. The statistician can grain an understanding of the contextual problem, and assess the appropriateness of data collection procedure to answer the underlying contextual questions. 

Beware of contextual problems that are "fishing expeditions": if you look hard enough, you'll almost always find something. But... that something may just be a coincidence of no meaningful consequence. Also, sometimes collaborators request a more complicated analysis is performed than is necessary. You may find that simple descriptive statistics are all that are really required. Even if a more complex analysis is necessary, simple descriptive statistics can provide a valuable supplement.

3. **Translate the problem into statistical terms**. This can be a challenging step and oftentimes requires creativity on the part of the statistician (in other words, not every problem fits into a neat little box like in an intro stats course!). Care must be taken at this step.

Connected to the above guidance is a good understanding of the **Data Collection** - It is vitally important to understand how the data was collected. Consider:

* Are the data observational or experimental? Are the data a "sample of convenience", or were they obtained via random sampling? Did the experiment include randomization? How the data were collected has a crucial impact on what conclusions can be meaningfully made. We discuss this topic more in Section \@ref(dataSources).
* Is there "non-response"? The data you don't see may be just as important as the data you do see! This aligns with the concern of whether a sample is representative of the intended target population.
* Are there missing values? This is a common problem that can be troublesome in some applications but not a concern in others. Data that are "missing" according to some pattern or systematic procedure are particularly troublesome.
* How are the data coded? In particular, how are the qualitative or categorical variables represented? This is important on the implementation of statistical methods in R (and other languages).
* What are the units of measurement? Sometimes data are collected or represented using far more digits than are necessary. Consider rounding if this will help with the interpretation. 
* Beware of data entry errors. This problem is all too common, and is almost a certainty in any real dataset of at least moderate size. Perform some data sanity checks.

## Introduction to R

Much like a chemist using test tubes & beakers, or a microbiologist using a microscope, the *laboratory instrument* for modern statistics is the computer. Many software packages exist to aid a modern statistician in the conducting of an analysis. This text uses the **R Language and Environment for Statistical Computing** [@R-base]. Many other tools exist and they can generally be lumped into two categories: *point and click* interfaces, and *coding* languages. The *point and click* tools tend to be easier to use but may have limited features & methods available. The *coding* languages are substantially more versatile, powerful and provide solutions that are reproducible, but may have a steeper learning curve. We recap a few of the common tools below.

* **Excel** - Microsoft Excel is a common *point and click* spreadsheet tool that includes many statistical methods. In fact, it is possible you used Excel in your Introductory Statistics course.
* **SPSS** - Also known as the Statistical Package for the Social Sciences, SPSS looks similar to Excel but has many more statistical tools included. SPSS includes the ability to code in Python.
* **Minitab** - Another *point and click* statistical tool. It's appearance is similar to Excel but, similar to SPSS, includes most common statistical methods.
* **JMP** - Pronounced as "jump". This tool includes a graphical interface (much like Excel) but also provides the user the ability to code. It is commonly used in the design of experiments and some intro stat courses.
* **SAS** - arguable the first statistical software that was widely used. SAS is a coding language that has powerful data handling abilities and includes implementations for many statistical methods. There is a graphical version known as Enterprise Miner.
* **Python** - A high-level general-purpose programming language, Python has become very popular in the Data Science industry due to its ease of use, the large community of support and add-ons (called modules) that implement data analysis methods.

**So why R??** - With so many tools available, you may ask, "why do we use R?" Simply put, R incorporates many of the features available in these other languages. Specifically,

* Similar to the other coding languages, R provides reproducible results!
* R is free!
* Installing a working version of R tends to be easier than Python.
* Similar to Python, R is fairly easy to learn.
   + But unlike in Python, performing statistical analysis in R is a bit more straightforward.
   + With Python, many Interactive Development Environments are popular, but RStudio is the standard for coding in R - this makes implementation in the classroom a bit easier since everyone uses the same tools.
* Similar to Python, R includes add-ons. In R we call them *packages* and we will make use of several.
   + Users can create their own packages to share with others!
* Similar to SAS, R has powerful data handling features and implementations for many statistical methods.



## Data frames

A statistical dataset in R is known as a data frame, or `data.frame`. 
It is a rectangular array of information (essentially a table or matrix or or spreadsheet). Rows of a data frame constitute different *observations* and the columns constitute different *variables* measured.

For example, imagine we record the name, class standing, height and age of all the students in a class: the first column will record each students' name, with the second column recording their standing (*e.g.*, junior, senior, etc...), the third column records the students' height and fourth column their age. Each row corresponds to a particular student (an observation) and each column a different property of that student (a variable).

### Built-in data

R packages come bundled with many pre-entered datasets for exploration purposes. For example, here is the data frame `stackloss` from the `datasets` package (which loads automatically upon starting R):

```{r ch1-1}
stackloss
```

This data frame contains 21 observations on 4 different variables. Note the listing of the variable names, this is known as the *header*.

Another famous dataset is the Edgar Anderson's Iris data and is included in R:

```{r ch1-2}
head(iris, n=10)
tail(iris, n=7)
```

Here the `head()` function prints the first few observations of the dataset, we specified the first 10 rows. The `tail()` functions prints out the last `n` rows (here specified to be 7). The size of the dataset is actually bigger and can be determined with the `dim` function,

```{r ch1-3}
dim(iris)
```

There are 150 observations on 5 variables (`Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width`, and `Species`).

Another useful function to quickly look at data is the `glimpse()` function in the R package `dplyr`, part of the `tidyverse` collection of packages [@R-tidyverse]. Here we apply the function to the built in dataset `esoph`, which is from a case-control study on esophageal cancer. 

```{r ch1-3a}
library(tidyverse)
glimpse(esoph)
```

There are 88 observations in this study with 5 recorded variables. 


### Types of Data{#dataTypesSection}

You'll notice in the above examples we see a mix of numeric and character values in the data frames. In R, every object has a type or *class*. To discover the type for a specific variable use the `class` function,

```{r ch1-4}
class(iris)
```

We see that the class of the object `iris` is a `data.frame` consisting of 150 rows and 5 columns. You should also note from above that the first four columns are numeric while the last column appears to be characters. We can explore the class of each variable by specifying the name of the column of the dataset using the `$` notation. First, let's get the list of variables in the dataset using the `names()` function.

```{r ch1-5}
names(iris)
class(iris$Sepal.Length)
class(iris$Species)
```

The `class` of `Sepal.Length` is numeric, which should not be too surprising given the values are numbers. The `class` of `Species` on the other hand is a `factor`. In R, a `factor` is **a categorical variable!** Here it happens to be labeled with the scientific species name, but R treats it as a **category**.

**Note**: The `factor` class can be quite important when performing Statistics. Specifically, when categories are coded as numbers it can be problematic. Depending if a variable is a `factor` or `numeric`, it can cause statistical methods to perform differently.

### Importing datasets into R

In practice you will often encounter datasets from other external sources. These are most commonly created or provided in formats that can be read in Excel. The best formats for importing dataset files into R is space-delimited (typically with the file extension `.txt`), tab-delimited text (either the file extension `.txt` or `.tsv`) or comma-separated values (file extension `.csv`). R can handle other formats, but these are the easiest with which to work.

There are multiple functions to import a data file in R. The tools we will outline here include many functions in the `readr` package [@R-readr], part of the `tidyverse`. We recommend reading through the Vignette by running the command `vignette("readr")` in the console.

#### `read_table()` -- for Space delimited files

One common tool is the `read_table()` command. Below are two examples:

1. Read a file directly from the web. The below code imports the file `univadmissions.txt` (details about the data are below) from the textbook data repository into the current R session, and assigns the name `uadata` to the data frame object:

```{r}
site <- "https://tjfisher19.github.io/introStatModeling/data/univadmissions.txt"
class(site)
```

```{r, eval=FALSE}
options(readr.show_col_types = FALSE)
uadata <- read_table(site)
head(uadata)
```

```{r, echo=FALSE, ch1-6}
options(readr.show_col_types = FALSE)
uadata <- read_table("docs/data/univadmissions.txt")
head(uadata)
```

A few notes on the above code. In the first line, we are creating a string of characters (that happens to be a web address) and *assigning* it to the R object `site` using the `<-` notation. You should note the `class` of the object `site` is `character`. The third line, `options(readr.show_col_types = FALSE)`, is *entirely optional* and tells the `readr` package to suppress some extra output from display. The key is we are applying the function `read_table()` with the file specified in the `site` variable. Lastly, the output of the function is being assigned to `uadata`, again using the `<-` notation. We will be using this same data again below.

2.  Another option is to download the file to your computer, then read it. You could go to the URL above and save the data file to your default working directory on your computer. After saving it into a folder (or directory), you can read the file with code similar to:

```{r, eval=FALSE}
uadata <- read_table("univadmissions.txt")
```

If you save the file to a directory on your computer other than the local directory (where your R code is stored), you must provide the path in the `read_table()` command:

```{r, eval=FALSE}
uadata <- read_table("C:/My Documents/My STA363 Files/univadmissions.txt")
```

The function `read_table()` automatically converts a space delimited text file to a data frame in R. 

#### `read_csv()` -- for Comma Separated Value files

Similar to the above examples, reading in a Comma Separated Value (CSV) file is very similar. This is becoming the standard format for transferring files. An example of some **non-executed code** is provided below:

```{r, eval=FALSE}
uadata_csv <- read_csv("c:/My Documents/My STA363 Files/univadmissions.csv")
```

**Note:** the file `univadmissions.csv` does not exist but this code has been included for example purposes.

#### `load()` -- Loading an existing R `data.frame`

If data has already been input and processed into R as a `data.frame`, it can be saved in a format designed for R using the `save()` function and then can be reloaded using `load()`. For example, first we may save the University Admidssion data

```{r, eval=FALSE}
save(uadata, file="uaDataInRformat.RData")
```

Then, sometime later, we may "load" that data into R

```{r, eval=FALSE}
load("uaDataInRformat.RData")
```

## Referencing data from inside a data frame

We saw above that we can access a specific variable within a dataset using the `$` notation. We can also access specific observations in the dataset based on rules we may wish to implement. As a working example we will consider the university admissions data we read in above.

**Example:** University admissions (originally in @KutnerText).

The director of admissions at a state university wanted to determine how accurately students' grade point averages at the end of their freshman year could be predicted by entrance test scores and high school class rank. The academic years covered 1996 through 2000. This example uses the `univadmissions.txt` dataset in the repository. The variables are as follows:

* `id` -- Identification number
* `gpa.endyr1` -- Grade point average following freshman year
* `hs.pct` -- High school class rank (as a percentile)
* `act` -- ACT entrance exam score
* `year` -- Calendar year the freshman entered university

This is the first dataset we input above in the `uadata` data frame.

The `tidyverse` package [@R-tidyverse] is actually a wrapper for many R packages. One of the key packages we use for data processing is the `dplyr` package [@R-dplyr]. It includes many functions that are similar to [SQL](https://www.tutorialspoint.com/sql/sql-overview.htm) (for those who know it). These tools tend to be very readable and easy to understand. First we will load the package `tidyverse` and then take a *glimpse* of the dataset:

```{r ch1-7, message=FALSE}
library(tidyverse)   # Load the necessary package
glimpse(uadata)      # Take a glimpse at the data
```

In the code above, you may note that everything to the right of a `#` is a *comment* -- this is, text included in code that does function or execute as code.  Comments are a useful tool to provide notes to use yourself about the code.

Key features when using the `tidyverse` is the pipe command, `|>` and *actions* on a `data.frame`. These *actions* are function calls and other sources may call them *verbs*, as you are performing some action on he data. We will introduce the function calls as needed.

**Note:** there are actually two different syntax for using the pipe in R, `|>` and `%>%`.  In the original version of this text, we used the latter `%>%` as the newer `|>` was not included in previous versions of R. At this point, `|>` is preferred but some examples may still use the older `%>%` notation.

As an example of some basic data processing, consider the case of extracting students with a GPA greater than 3.9:

* Take the `data.frame` `uadata` (from above) and pipe it (send it) to a function (or action), in this case the *filter* function.
* Then `filter` it based on some criteria, say `gpa.endyr1 > 3.9`.

```{r ch1-8, warning=FALSE, error=FALSE}
uadata |>
  filter(gpa.endyr1 > 3.9)
```

Another useful `dplyr` command for choosing specific variables is the `select` function. Below we select the ACT scores for students who have a first year GPA greater than 3.9 and with a graduating year of 1998.

```{r ch1-9}
uadata |>
  filter(gpa.endyr1 > 3.9, 
         year==1998) |>
  select(act)
```

Note in the above, we take the `uadata` and pipe it to a `filter`, which selects only those observations with `gpa.endyr1` greater than 3.9 and `year` equal (`==`) to 1998, then we pipe that subset of the data to the `select` function that returns the variable `act`.

**IMPORTANT NOTE** - R is **an *evolving* language** and as such occasionally has some wonky behavior. In fact, there are multiple versions of both the `filter` and `select` functions in R. If you ever use one and receive a strange error (`comparison is possible only for atomic and list types` or an `unused argument` error), first check for a typo. If the error still occurs it is likely the case that R is trying to use a different version of the `filter` or `select`. You can always force R to use the `dplyr` version by specifying the package as in the below example:

```{r eval=FALSE}
uadata |>
  dplyr::filter(gpa.endyr1 > 3.9, 
                year==1998) |>
  dplyr::select(act)
```

## Missing values and computer arithmetic in R

Sometimes data from an individual is lost or not available, these are known as missing values. R indicates such occurrences using the value `NA`. Missing values often need to be addressed because simply ignoring them when performing certain functions will result in no useful output. 

As an example, we cannot find the mean of the elements of a data vector when some are set to `NA`. Consider the simple case of a vector of 6 values with one `NA`.

```{r ch1-10}
x <- c(14, 18, 12, NA, 22, 16)
class(x)
mean(x)
```

Here, the object `x` is a *vector* of class `numeric`. We will not be working with vectors much in this class (although a column in a data frame can be considered a vector) but we do so here to demonstrate functionality.

Since one of the observations is missing, `NA`, the mean of the vector is also "not available" or `NA`.\
This is not incorrect: logically speaking there is no way to know the mean of these observations when one of them is missing. However, in many cases we may still wish to calculate the mean of the non-missing observations.

**Many R functions have a `na.rm=` logical argument**, which is set to `FALSE` by default. It basically asks if you want to skip, or remove, the missing values before applying the function. It can be set to `TRUE` in order to remove the `NA` terms. For example,

```{r ch1-11}
mean(x, na.rm=TRUE)    # mean of the non-missing elements
```

The number system used by R has the following ways of accommodating results of calculations on missing values or extreme values:

* `NA`: **Not Available**. Any data value, numeric or not, can be `NA`. This is what you use for missing data. Always use `NA` for this purpose.

* `NaN`: **Not a Number**. This is a special value that only numeric variables can take. It is the result of an undefined operation such as 0/0.

* `Inf`: **Infinity**. Numeric variables can also take the values `-Inf` and `Inf`. These are produced by the low level arithmetic of all modern computers by operations such as 1/0. (You should not think of these as values as real infinities, but rather the result of the correct calculation if the computer could handle extremely large (or extremely small, near 0) numbers; that is, results that are larger than the largest numbers the computer can hold (about $10^{300}$)).

**Scientific notation**. This is a shorthand way of displaying very small or very large numbers. For example, if R displays `3e-8` as a value, it means $3\times 10^{-8} = 0.00000003$. Similarly, `1.2e4` is $1.2\times 10^4 = 12000$. 


## Exploratory Data Analysis (EDA)

The material presented in this section is typically called *descriptive statistics* in Intro Statistics books. This is an important step that should always be performed prior to a formal analysis. It looks simple but it is vital to conducting a meaningful analysis. EDA is comprised of:

* Numerical summaries 
   + Means & medians (measures of centrality)
   + Standard deviations, range, Interquartile Range (IQR) (measures of spread)
   + Percentiles/quantiles & Five-number summaries (assessment of distributions)
   + Correlations (measures of association)
* Graphical summaries
   + One variable: boxplots, histograms, density plots, etc.
   + Two variables: scatterplots, side-by-side boxplots, overlaid density plots
   + Many variables: scatterplot matrices, interactive graphics

We look for outliers, data-entry errors, skewness or unusual distributions using EDA. 
We may ask ourselves: "are the data distributed as you expect?" 
Getting data into a form suitable for analysis by cleaning out mistakes and aberrations is often time consuming. 
It often takes more time than the data analysis itself! 
In this course, data will usually be ready to analyze but we will occasionally intermix messy data. 
You should realize that in practice it is rare to receive *clean* data.

### Numeric Summaries

#### Measures of centrality and quantiles

Common descriptive measures of centrality and quantiles from a sample include

* Sample mean -- typically denoted by $\bar{x}$, this is implemented in the function `mean()`
* Sample median -- this is implemented in the function `median()`
* Quantiles -- Quantiles, or sample percentiles, can be found using the `quantile()` function.
   + Three common quantiles are $Q_1$ (the sample $25^\mathrm{th}$ percentile, or first quartile), $Q_2$ (the $50^\mathrm{th}$ percentile, second quartile or *median*) and $Q_3$ (the sample $75^\mathrm{th}$ percentile or third quartile).
* Min and Max -- the Minimum and Maximum values of a sample can be found using the `min()` or `max()` functions.

Each of these measures are covered in a typical introductory statistics course. We refer you to that material for more details.

#### Measuring variability{#variabilityDiscussion}

Earlier we defined statistics as the 'study of variation'. The truth is that everything naturally varies:

* If you measure the same characteristic on two different individuals, you would likely get two different values (subject variability).
   + *e.g.*, the pulse rate of two individuals is likely to be different.
* If you measure the same characteristic twice on the same individual, you may get two different values (measurement error).
   + *e.g.*, if two different people manually measured someone's pulse rate, you may get different results.
* If you measure the same characteristic on the same individual but at different times, you would potentially get two different values (temporal variability).
   + *e.g.*, measuring someone's pulse rate while resting compared to while performing exercise sometime later.

Because almost everything varies, simply determining that a variable *varies* is not very interesting. 
The goal is to find a way to discriminate between variation that is scientifically interesting and variation that reflects background fluctuations that occur naturally. 
This is what the science of statistics does!

Having some intuition about the amount of variation that we would expect to occur just by chance alone when nothing scientifically interesting is key. 
If we observe bigger differences than we would expect by chance alone, we may say the result is statistically significant. 
If instead we observe differences no larger than we would expect by chance alone, we say the result is not statistically significant.

Although often covered in introductory statistics courses, due to importance we review measuring the variance of a sample.

**Variance**

A *variance* is a numeric descriptor of the variation of a quantity. 
Variances will serve as the foundation for much of what we use in making statistical inferences in this course. 
In fact, a standard method of data analysis for many of the problems we will encounter is known as analysis of variance (shorthanded ANOVA).

You were exposed to variance in your introductory statistics course. 
What is crucial now is that you understand what you are finding the variance of; *i.e.*, what is the "quantity" we referred to in the preceding paragraph? 
Consider the following example:

* *Example*. Suppose you collect a simple random sample of fasting blood glucose measurements from a population of $n$ diabetic patients, where $n$ is the standard notation for the size of a sample. The measurements, or observations, are denoted $x_1$, $x_2$, $\ldots$, $x_n$. What are some of the different "variances" that one might encounter in this scenario?

    1. The $n$ sampled measurements vary around their sample mean $\bar{x}$. This is called the sample variance (traditionally denoted $s^2$). The formula for $s^2$ is provided below. In R, we compute this with the function `var()`.
    2. All fasting blood glucose values in the population vary around the true mean of the population (typically denoted with $\mu$, the lowercase Greek 'mu'). This is called the population variance (usually denoted $\sigma^2$ where $\sigma$ is the lowercase Greek 'sigma'). Note that $\sigma^2$ cannot be calculated without a census of the entire population, so $s^2$ serves as an estimate of $\sigma^2$.
    3. $\bar{x}$ is only a sample estimate of $\mu$, so we should expect that it will vary depending on which $n$ individual diabetic patients are sampled (*i.e.*, with a different sample we will calculate a different sample mean). This is called the variance of $\bar{x}$, and is denoted $\sigma_{\bar{x}}^2$. This variance, unlike the sample or population variance, describes how sample estimates of $\mu$ vary around $\mu$ itself. We will need to estimate $\sigma_{\bar{x}}^2$, since it cannot be determined without a full survey of the population. The estimate of $\sigma_{\bar{x}}^2$ is denoted $s_{\bar{x}}^2$.

The variances cited above are each different measures. Even more measures of variability exist; for example, each different possible random sample of $n$ patients will produce a different sample variance $s^2$. Since these would all be estimates of the population variance $\sigma^2$, we could even find the variance of the sample variances!

Some formulas:

$$\textbf{Sample Variance:}~~~ s^2 = \frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2, \textrm{ for sample size } n$$

$$\textbf{Estimated Variance of}~ \bar{x}:~~~ s^2_{\bar{x}} = \frac{s^2}{n}$$

Three important notes:

1. The general form of any variance is a sum of squared deviations, divided by a quantity related to the number of independent elements in the sum known as the **degrees of freedom**. So in general,

$$\textrm{Variance} = \frac{\textrm{sum of square deviations}}{\textrm{degrees of freedom}} = \frac{SS}{\textrm{df}}$$

2. The formula for the population variance $\sigma^2$ is omitted since it is rarely used and requires the $\mu$ to be a known value. In practice, this can only be calculated with a population census.

3. When considering how a statistic (like $\bar{x}$) varies around the parameter it is estimating ($\mu$), it makes sense that the more data you have, the better it should perform as an estimator. This is precisely why the variance of $\bar{x}$ is inversely related to the sample size $n$. **As** $n$ **increases**, the variance of the estimate decreases; *i.e.*, the estimate will be closer to the thing it is estimating. This phenomenon is known as the law of large numbers and is typically discussed in Introductory Statistics courses.

**Standard deviations and standard errors**

While variances are the usual items of interest for statistical testing, variances are always in units squared (look at the equation for $s^2$ above). 
Often we will find that a description of variation in the original units of the quantity being studied is preferred, so we take their square roots and are back to the original units. The square root of a variance is called a **standard deviation** (SD). 
The square root of the variance of an estimate is called the **standard error** (SE) of the estimate:

$$\textbf{Sample Standard Deviation:}~~~ s = \sqrt{s^2} = \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i - \bar{x})^2}, ~~~\textrm{ for sample size } n$$

$$\textbf{Standard error of}~ \bar{x}:~~~ s_{\bar{x}} = \sqrt{s^2_{\bar{x}}} = \sqrt{\frac{s^2}{n}} = \frac{s}{\sqrt{n}}$$

### Numeric Summaries in R{#numericSummaries}

We recommend using `dplyr` for data handling and summary statistics. We can pipe (`|>`) our data frame into a function to calculate numeric summaries, this function is `summarize()`.

```{r ch1-12}
uadata |>
  pivot_longer(everything(), names_to="Variable", values_to="value") |> 
  group_by(Variable) |>
  summarize(across(everything(), 
                   list(Mean=mean, SD=sd, Min=min, Median=median, Max=max) ) )
```

For a brief moment ignore the code (we described below). Presented here are the mean, standard deviation, the minimum, median and maximum of each numeric variable in the University Admissions data frame. For example, the mean freshman year-end GPA over the period 1996-2000 was 2.98; the median over the same period was 3.05, suggesting the GPA distribution is slightly skewed left (not surprising). The range of the GPA distribution is from 0.51 to 4. The median ACT score for entering freshman was 25, and and the range of ACT scores was from 13 to 35.

#### Coding Notes

In `tidyverse` there is package called `tidyr` [@R-tidyr] which is mainly comprised of two main functions: `pivot_longer()` and `pivot_wider()`. 
The purpose of the `pivot_longer()` function is to collect variables in a dataset and create a *tall* (or *longer*) dataset (this is typically known as going from wide-to-tall or wide-to-long format). 
We specify the variables (or columns) we want to pivot (here we pivot on everything, thus `everything()` in the code), the name of the new variable (here called `Variable`) storing the column names, and the variable name that will store the values (here called `value`). 
The `pivot_wider()` function essentially does the opposite, turning a tall data frame into a wide format.

In `dplyr` resides the function `group_by()`. 
This function essentially does some *internal organization* of our data. 
Specifically, for each unique `Variable` name, it would group the data together. 
So, all the `gpa.endyr1` values will be considered a group, and separate from all the `hs.pct` and `act` *groups*. 
This function is particularly useful when coupled with `summarize()` (such as in the example above) because summary statistics will be calculated for each group.

The function `summarize()` tells R to calculate summary statistics across rows (or grouped rows) within a `data.frame`.  In the above example, we compute a list of summary statistics (hence, `list(Mean=mean, SD=sd, Min=min, Median=median, Max=max)`) on all the non-grouped variables (`everything()`). We will see more examples below.

The code above introduces a few keys points to know:

* The code is spread across multiple lines with most lines ending with the piping operator `|>`. This is done for readability (and recommended by all users of R). When performing an analysis, you often will revisit the code at a later time. Readable code is paramount to reproducibility.
* The `tidyverse` structure of data processing is meant to be highly readable. Look back at the code and consider the following:
   + We get the university admissions data, and then send it to  (`uadata |>`)
   + the function `pivot_longer()` that takes all columns, or variables (`everything()`) in the data set and stacks them on top of one another where a new variable called `Variable` designates the variable from the wide version of the data, and the new variable `Values` provides the corresponding value, and then send it to (`|>`)
   + a function that internally organizes the dataset, treating all `Values` with the same `Variable` value as a group (`group_by(Variable)`), and then send it to (`|>`)
   + a function that will calculate summary statistics (`summarize()`). This includes the mean, standard deviation, minimum, median and maximum of the `Values` for each `Variable` value since the data had been grouped.
   
Also note:

* Since we applied the command to the entire data frame (`everything()`), we also see numeric summaries of the variable `id`, even though it is meaningless to do. (note: just because a computer outputs it, that doesn't mean it's useful)
* If you look carefully, you'll note that `year` is treated as a number as well (*e.g.*, 1996, 1997, etc.). However, that doesn't necessarily mean it should be treated as a numeric variable! It is meaningless to find numeric summaries, such as means or medians, for the `year` variable because it is qualitative (categorical) variable in the context of these data (again, just because a computer outputs it, that does not mean it is contextually correct).

These are common mistake to make. When R encounters a variable whose values are entered as numbers in a data frame, R will treat the variable as if it were a numeric variable (whether it is contextually or not). 
If R encounters a variable whose values are character strings, R will treat the variable as a categorical `factor`. 
But, in a case like `year` or `id` above, you have to explicitly tell R that the variable is really a `factor` since it was coded with numbers. 
To do this, we will `mutate` the variables `id` and `year` and turn them into factors.

```{r ch1-13}
uadata <- uadata |>
  mutate(id=factor(id),
         year=factor(year))
```

Here we tell R to take the `uadata` data, pipe it to the `mutate` function (which allows us to create and/or replace variables by changing/mutating them). We then assign the variables `id` and `year` as versions of themselves that are being treated as *factors* (or categories).

We can then check the type of each variable by *summarizing* over *all* the variables using the `class` function.

```{r ch1-14}
uadata |>
  summarize(across(everything(), class))
```

Now calculate the summary statistics for only numeric variables (*i.e.*, the use of the `select(where(is.numeric))` function).

```{r ch1-15}
uadata |>
  select(where(is.numeric)) |>
  pivot_longer(everything(), names_to="Variable", values_to="value") |> 
  group_by(Variable) |>
  summarize(across(everything(), 
                   list(Mean=mean, SD=sd, Min=min, Median=median, Max=max) ) )
```

When a dataset includes categorical variables, the main numeric measure to consider is the count (or proportions) of each category. To do this in R, we tell it to treat each category as a *group* and then count within the group. We will `group_by` the `year` and then within that year, count the number of observations (`summarize(Count=n() )`. We also add a new variable (using `mutate`) that is the proportion.

```{r ch1-16}
uadata |>
  group_by(year) |>
  summarize(Count=n() ) |>
  mutate(Prop = Count/sum(Count))
```

Note the power of using the `tidyverse` tools [@R-tidyverse]. 
We are only *scratching the surface* of what can be done here. 
For more information, we recommend looking into the *Wrangling* chapters in the book [R for Data Science](https://r4ds.had.co.nz/index.html) [@Wickham2017R].

### Graphical Summaries

We begin a graphical summary with a matrix (or table) of univariate and bivariate plots using the `ggpairs()` function in the `GGally` package [@R-GGally]. The results can be seen in Figure \@ref(fig:ch1-17).

```{r ch1-17, message=FALSE, fig.align='center', fig.cap='Pairs plot showing the relationship between the variables `gpa.endy1`, `hs.pct`, `act` and `year`.', fig.asp=1, out.width='70%'}
library(GGally)
ggpairs(uadata, columns=2:5)
```

#### Coding note

Here we specify to only plot using the variables in columns 2, 3, 4 and 5 (the `2:5`). We do this because the `id` variable in `uadata` is an identifier (column 1) and unique identifies each observations -- it is not a variable of statistical interest -- we do not want to try and plot it.

Note that by default the function provides a collection of univariate and bivariate plots (Figure \@ref(fig:ch1-17)). 
We recap each of the plots provided (further details on each plot type is provided below).

* On the diagonal of the matrix we see univariate plots -- density plots for the numeric variables and a bargraph of frequencies for categorical variable `year.
* In the upper right triangle of the matrix we see numeric measures for the amount of correlation between each of the numeric variables (Pearson Correlation coefficients, typically denoted with $r$). In the case of a numeric variable matched with a categorical variable, side-by-side boxplots are provided.
* In the bottom left triangle of the matrix scatterplots are provided to show the relationship between numeric variables. We also see histograms for each of the numerical variable stratified by the categorical variable `year`. Now we will discuss each of the elements in more detail.

### Distribution of Univariate Variables

The `ggpairs` function provides a look at the distributional shape of each variable using what is known as a *density plot* (this can be consider a more-modern, and arguably less subjective, histogram). The `ggpairs` function builds off a function known as `ggplot()` from the `ggplot2` package [@R-ggplot2] which is included in `tidyverse` [@R-tidyverse].

We can build density plots directly using the `geom_density()` function as seen in the code below. The resulting density plot is in Figure \@ref(fig:ch1-18).

```{r ch1-18, fig.align='center', fig.cap='Density plot showing left-skewed distribution for the `gpa.endyr1` variable', out.width='60%', fig.asp=0.55}
ggplot(uadata) + 
  geom_density(aes(x=gpa.endyr1))
```

#### Coding Notes

Note the code is similar to the `|>` operator but with a `+`. 
The trick to working with plotting on the computer is to think in terms of *layers*. 
The first line `ggplot(uadata)` sets up a plot object based on the dataset `uadata`. 
The `+` says to add a layer, this layer is a density plot (`geom_density`) where the aesthetic values include an $x$-axis variable matching that of `gpa.endyr`.

To generate even more complex distributional plots, first we need to perform some data processing. 
Above we saw `pivot_longer()`. 
The purpose of the `pivot_longer()` function is to collect variables in a dataset and create a *tall* (or *longer*, wide-to-long) dataset. 
We specify the variables (or columns) we want to pivot (the `c(gpa.endyr1, hs.pct, act)` part in the below code), the name of the new variable (here called `var_names`) storing the column names, and the variable name that will store the values (here called `values`). 

```{r ch1-19}
uadata.tall <- uadata |>
  pivot_longer(c(gpa.endyr1, hs.pct, act), 
               names_to="var_names", values_to="values" )
head(uadata)
head(uadata.tall)
dim(uadata)
dim(uadata.tall)
```

Take a minute and step through the above code to understand it. 
The data is now in a *tall*, or *long*, format. 
We can now utilize the `ggplot2` package for more advanced plotting. 
Here we want to draw a density plot for the `values` in the tall, or long, version of the dataset, called `uadata.tall`.
But we want a different plot for each `var_names` *variable* (that is, `act`, `gpa.endyr1`, and `hs.pct`) so we add a `facet_grid()` layer. 
Figure \@ref(fig:ch1-20) provides the result for the below code.

```{r ch1-20, fig.align='center', fig.cap='Side-by-side density plots showing the relative distributions of `act`, `gpa.endyr1` and `hs.pct`.', , out.width='70%', fig.asp=0.50}
ggplot(uadata.tall) +
  geom_density(aes(x=values)) + 
  facet_grid(.~var_names, scales="free_x") + 
  theme_bw()
```

#### Coding Notes

The `facet_grid` option tells the plot to draw each density plot for each observed value in `var_names`. 
The *facet* feature is similar to the `group_by` statement when calculating summary statistics.
The `theme_bw()` specifies the plotting style (there are many others but we will typically use `theme_bw()`, `theme_minimal()` or the default behavior which corresponds to `theme_grey()`). 
The `free_x` option tells the plot that each $x$-axis is on a different scale. 
Feel free to modify the code to see how the plot changes!

We can see that GPA and high school percentile rankings are skewed left, and that ACT scores are reasonably symmetrically distributed.

Alternatively, we could build histograms to make the comparison.

```{r ch1-20a, fig.align='center', fig.cap='Stacked histogram showing the relative distributions of `act`, `gpa.endyr1` and `hs.pct`.', , out.width='60%', fig.asp=0.60}
ggplot(uadata.tall) +
  geom_histogram(aes(x=values), binwidth = 2.5) + 
  facet_grid(var_names~., scales="free_y") + 
  theme_bw()
```

Note the similarities and difference in Figures \@ref(fig:ch1-20) and \@ref(fig:ch1-20a). 
Depending on whether we facet on the $y$-axis or $x$-axis we can get different appearances.
Also note that the code for plotting a histogram is similar to that of a density plot (`geom_histogram` vs `geom_density`).

**Bar Graphs**

The best visual tool for simple counts is the Bar Graph (**note:** you should never use a pie chart, see [here for an explanation](http://www.businessinsider.com/pie-charts-are-the-worst-2013-6) ). 
The below code generates a simple (and fairly boring) bar graph (Figure \@ref(fig:ch1-21)) displaying the number of observations for each `year` in the `uadata`.

```{r ch1-21, fig.align='center', fig.cap='Bar graph displaying the number of observations in the `uadata` per `year`.', out.width='60%', fig.asp=0.55}
ggplot(uadata) + 
  geom_bar(aes(x=year) )
```

We see here that nothing too interesting is happening. A fairly uniform distribution of counts is seen across years.



### Descriptive statistics and visualizations by levels of a factor variable

While the above summaries are informative, they overlook the fact that we have several years of data and that it may be more interesting to see how things might be changing year-to-year. This is why treating `year` as a factor is important, because it enables us to split out descriptions of the numeric variables by levels of the factor.

One of the simplest and best visualization tool for comparing distributions is the side-by-side boxplot, which displays information about quartiles ($Q_1$, $Q_2$ and $Q_3$) and potential outliers. The below code generates Figure \@ref(fig:ch1-22).

```{r ch1-22, fig.align='center', fig.cap='Side-by-side Box-whiskers plots showing the distribution of ACT scores in the `uadata` per year.', out.width='60%', fig.asp=0.55}
ggplot(uadata) +
  geom_boxplot(aes(x=act, y=year) ) +
  labs(title="ACT scores of Incoming Freshmen",
       x="Composite ACT Score",
       y="Year") + 
  theme_classic()
```

The code is similar to the examples above except now we call the `geom_boxplot` function. 
Also note we are using `theme_classic()` which has some subtle differences compared to `theme_bw()` used above. 
Lastly, we include proper labels and titles on our plot here for a more professional appearance.

Next we construct side-by-side boxplots by year for each of the three numeric variables in the data frame. 
We utilize the taller or longer format of the admissions data in `uadata.tall`. 
Two versions of this plot are provided in Figure \@ref(fig:ch1-23) and Figure \@ref(fig:ch1-23a).

```{r ch1-23, fig.align='center', fig.cap='Side-by-side Box-whiskers plots showing the distributions of the variables `act`, `gpa.endyr1` and `hs.pct` for each year in the `uadata` dataset.', out.width='70%'}
# Code for Figure 1.6
ggplot(uadata.tall) + 
  geom_boxplot(aes(x=values, y=year) ) +
  facet_grid(.~var_names, scales="free_x") + 
  theme_dark()
```

Note another theme option, `theme_dark()` (obviously you will NOT want to use this theme if planning to print the document). 
A more aesthetically pleasing version of the plot (with proper labeling) is generated with the code below and can be seen in Figure \@ref(fig:ch1-23a).

```{r ch1-23a, fig.align='center', fig.cap='Side-by-side Box-whiskers plots showing the distributions of the variables `act`, `gpa.endyr1` and `hs.pct` for each year in the `uadata` dataset.', out.width='70%'}
# Code for Figure 1.7
ggplot(uadata.tall) + 
  geom_boxplot(aes(x=values, y=year) ) +
  facet_grid(.~var_names, scales="free_x") + 
  labs(y="Year", x="", title="Distribution of ACT, GPA and HS Rank by Year") +
  theme_minimal()
```

This is a nice compact visual description of the changes in patterns of response of these three variables over time.\
All three variables appear pretty stable over the years; the left skews in GPA and high school percentile rankings are still evident here (note the detection of some low outliers on these two variables).

We now revisit calculating summary statistics. 
Extensive grouped numeric summary statistics are available by using the `group_by()` and `summarize()` functions in `tidyverse`.
Recall that the `group_by` function tells R to treat each value of the specified variable (in this case, called `var_names`) as a collection of similar options, then to perform the `summarize` function on each value in the grouped variables.
The below example is a more detailed version of the summary table from above.

```{r ch1-24}
library(knitr)

summary.table <- uadata.tall |>
  group_by(var_names) |>
  summarize(Mean=mean(values),
            StDev=sd(values),
            Min=min(values),
            Q2=quantile(values, prob=0.25),
            Median=median(values),
            Q3=quantile(values, prob=0.75),
            Max=max(values) )

kable(summary.table, digits=3) |>
  kable_styling(full_width=FALSE)
```

Here, we first pipe the `uadata.tall` into a `group_by()` function grouping by the `var_names` variable. 
Then we take that output and pipe it to the `summarize` function which allows us to calculate various summary statistics for each of the groups. 
In the above we calculate the mean and standard deviation of each group, along with the 5-number summary. 
Lastly, we wrap the constructed table (its technically a data frame) in the `kable()` function from the `knitr` package [@R-knitr] (where `kable` means 'knitr' 'table') for a nice display.
We pipe that output to the `kable_styling()` function in the `kableExtra` package [@R-kableExtra] for some additional formatting options (here, we condence the width of the table).

In the next example we group by both the `year` and the variables of interest.

```{r ch1-25, results=TRUE}
summary.table2 <- uadata.tall |>
  group_by(var_names, year) |>
  summarize(Mean=mean(values),
            StDev=sd(values))
kable(summary.table2, digits=3) |>
  kable_styling(full_width=FALSE)
```

### Descriptive statistics and visualizations for two numeric variables

For a pair of numeric variables the amount of linear association between the variables can be measured with the Pearson correlation.
The Pearson correlation assesses the *strength* and *direction* of any linear relationship existing between two numeric variables.
The correlation coefficients between the GPA, ACT and HS rank variables are provided Figure \@ref(fig:ch1-17).

Recall from introductory statistics that the Pearson correlation $r$ between two variables $x$ and $y$ is given by the formula 
$$r = \frac{1}{n-1}\sum_{i=1}^n\left(\frac{x_i - \bar{x}}{s_x}\right)\left(\frac{y_i - \bar{y}}{s_y}\right).$$
$r$ is a unitless measure with the property that $-1 \leq r \leq 1$.\
Values of $r$ near $\pm 1$ indicate a near perfect (positive or negative) linear relationship between $x$ and $y$. 
A frequently used rule of thumb is that $|r| > 0.80$ or so qualifies as a strong linear relationship. 
We can calculate the Pearson correlation with the `cor()` function in R. 
When applied to a `data.frame`, or `matrix`, it displays a correlation matrix (correlation of all variable pairs). 
In the below code we specify to only calculate the correlation based on the 3 numeric variables.

```{r ch1-25_cor}
cor(uadata |> select(gpa.endyr1, hs.pct, act))
```

All pairwise correlations in this example are of modest magnitude ($r \approx 0.4$) (see Figure \@ref(fig:ch1-17) or the above correlation matrix).

To visualize the relationship between numeric variables, we typically use a scatterplot.
A scatterplot is a simple bivariate plot where dots, or points, are plotted at each ($x,y$) pair on the Cartesian plane.
The code below uses `ggplot`() to build a scatterplot using `geom_point()`.

```{r ch1-25-scatter, fig.align='center', fig.cap='Scatterplot showing the relationship between the variables `act` and `gpa.endyr1` in the `uadata` dataset.', out.width='60%', fig.asp=0.55}
ggplot(uadata) + 
  geom_point(aes(x=act, y=gpa.endyr1) ) +
  labs(x="Composite ACT Score",
       y="Grade point average following freshman year") +
  theme_minimal()
```

You should note in Figure \@ref(fig:ch1-25-scatter) the overall positive relationship (higher ACT scores tend to correspond to higher GPAs and corresponds to the Pearson correlation value of $0.366$). 
We also note that due to the rounding of the ACT scores (always a whole number) the resulting plot has vertical striations.

We attempt to visually address the striations with *jittering* (adding some visual random perturbations) in Figure \@ref(fig:ch1-25-jitter) with the code below using `geom_jitter()`. 
There, we specify no jittering on the vertical axis (thus, `height=0`) and let `ggplot2` use the default behavior horizontally.

```{r ch1-25-jitter, fig.align='center', fig.cap='Jittered scatterplot showing the relationship between the variables `act` and `gpa.endyr1` in the `uadata` dataset.', out.width='60%', fig.asp=0.55}
ggplot(uadata) + 
  geom_jitter(aes(x=act, y=gpa.endyr1), height=0 ) +
  labs(x="Composite ACT Score",
       y="Grade point average following freshman year") +
  theme_minimal()
```

Note, with jittering we are only adding *randomness* to the **plot**, we are not changing the recorded values in the `data.frame`.

## Sampling distributions: describing how a statistic varies

As described in the discussion on variability above, each of the numeric quantities calculated has some associated uncertainty.
It is not enough to simply know the variance or standard deviation of an estimate.
That only provides one measure of the inconsistency in an estimate's value from sample to sample of a given size $n$.

If we wish to make a confident conclusion about an unknown population parameter using a sample statistic (*e.g.*, $\bar{x}$), we also need to know something about the general pattern of behavior that statistic will take if we had observed it over many, many different random samples.
This long-run behavior pattern is described via the probability distribution of the statistic, which is known as the **sampling distribution** of the statistic.

It is important to remember that we usually only collect one sample, so only one sample estimate will be available in practice.
However, knowledge of the sampling distribution of our statistic or estimate is important because it will enable us to assess the reliability or confidence we can place in any generalizations we make about the population using our estimate.

**Two useful sampling distributions:** Normal and $t$

You should be very familiar with Normal distributions and $t$-distributions from your introductory statistics course.
In particular, both serve as useful sampling distributions to describe the behavior of the sample mean $\bar{x}$ as an estimator for a population mean $\mu$. We briefly review them here.

**Normal distributions.** 

A random variable $X$ that is normally distributed is one that has a symmetric bell-shaped density curve.
A normal density curve has two parameters: its mean $\mu$ and its variance $\sigma^2$.
Notationally, we indicate that a random variable $X$ is normal by writing $X \sim N(\mu, \sigma^2)$.

The normal distribution serves as a good sampling distribution model for the sample mean $\bar{x}$ because of the powerful Central Limit Theorem, which states that if $n$ is sufficiently large, the sampling distribution of $\bar{x}$ will be approximately normal, regardless of the shape of the distribution of the original parent population from which you collect your sample (when making a few assumptions).

**t distributions.** 

The formula given by 
$$t_0 = \frac{\bar{x}-\mu}{{SE}_{\bar{x}}}$$
counts the number of standard error units between the true value of $\mu$ and its sample estimate $\bar{x}$; it can be thought of as the standardized distance between a sample mean estimate and the true mean value.
If the original population was normally distributed, then the quantity $t_0$ will follow the Students' $t$-distribution with $n - 1$ degrees of freedom. 
The *degrees of freedom* is the only parameter of a $t$-distribution.\
Notationally, we indicate this by writing $t_0 \sim t(\textrm{df})$ or $t_0 \sim t(\nu)$ where $\nu$ is a lowercase Greek 'nu' (pronounced 'new') and is the notation for degrees of freedom.

$t$-distributions look like the standard normal distribution, $N(0, 1)$: they are both centered at 0, are symmetric and bell-shaped (see below).
The key difference is due to the fact that the quantity $t$ includes two **sample summaries** in its construction: the sample mean $\bar{x}$, and $SE_{\bar{x}}$, which incorporates the sample standard deviation $s$.
Because of this, $t$ is more variable than $N(0, 1)$ (more room for noise).
However, as $n$ (and hence degrees of freedom) increases, $s$ becomes a better estimate of the true population standard deviation, so the distinction ultimately vanishes with larger samples. 
This can all be visualized in Figure \@ref(fig:ch1-26) where the degress of freedom increase from 2 to 10 and you see the $t$-distribution approaches the Normal bell curve.

```{r ch1-26, echo=FALSE, fig.align='center', fig.cap='Overlayed distribution functions showing the relationship between a standard normal distribution and that of a $t$ distribution with 3, 5 and 10 degrees of freedom. As the degrees of freedom increases, the $t$ distribution becomes more *normal*', out.width='75%', fig.asp=0.55}
x.val <- seq(-3.5, 3.5, 0.01)
y.norm <- dnorm(x.val)
y.t2 <- dt(x.val, df=2)
y.t5 <- dt(x.val, df=5)
y.t10 <- dt(x.val, df=10)
my.data <- data.frame(x.val=x.val, Normal=y.norm, t2=y.t2, t5=y.t5, t10=y.t10)
my.data.tall <- my.data |>
  gather(key="key", value="value", -x.val) |>
  mutate(key=factor(key, levels=c("Normal", "t2", "t5", "t10")))
ggplot(my.data.tall) +
  geom_line(aes(x=x.val, y=value, col=key, linetype=key), linewidth=0.75) +
  scale_colour_manual(labels=c("Normal", "t(df=2)", "t(df=5)", "t(df=10)"), values=c("black", "navy", "dodgerblue",  "skyblue2")) + 
  scale_linetype_manual(labels=c("Normal", "t(df=2)", "t(df=5)", "t(df=10)"), 
                        values=c("solid","dashed","longdash","F1")) + 
  scale_x_continuous(breaks=seq(-3,3,1), minor_breaks = seq(-3.5,3.5, 1) ) +
  labs(x="", y="Density Function value", linetype="", color="") + 
  theme_classic() +
  theme(legend.position = "inside",
        legend.position.inside = c(0.8, 0.9),
        legend.key.width=unit(2, "cm"))
```

The $t$-distribution is usually the relevant sampling distribution when we use a sample mean $\bar{x}$ to make inferences about the corresponding population mean $\mu$. 
Later in the course, the need to make more complicated inferences will require us to develop different sampling distribution models to facilitate the statistics involved. These will be presented as needed.

## Two-sample inference{#twoSampleInference}

A key concept from your introductory statistics course that builds off sampling distribution theory is the comparison of the mean from two populations based on the sample means using a "two-sample $t$-test". 
Consider testing 
$$H_0:\ \mu_A = \mu_B ~~~\textrm{versus}~~~ H_A:\ \mu_A <\neq> \mu_B$$
where the notation $<\neq>$ represents one of the common alternative hypotheses from your introductory statistics course (less than $<$, not equal to $\neq$, or greater than $>$) and, $\mu_A$ is the population mean from population $A$ and $\mu_B$ is the population mean from population $B$. 
We statistically test this hypothesis via 
$$t_0 = \frac{\bar{x}_A - \bar{x}_B}{{SE}_{\bar{x}_A-\bar{x}_B}}$$
where ${SE}_{\bar{x}_A-\bar{x}_B}$ is the standard error of the difference of sample means $\bar{x}_A$ and $\bar{x}_B$. 
It can be estimated in one of two ways (depending if we assume the population variances are equal or not) and we reference your intro stat class for details (you may recall, pooled versus non-pooled variance). 
In R, we can perform this test, and construct the associated $(1-\alpha)100\%$ confidence interval for $\mu_A-\mu_B$, 
$$\bar{x}_A - \bar{x}_B \pm t_{\alpha/2} {SE}_{\bar{x}_A-\bar{x}_B}$$ 
using the function `t.test()`. 
Here $t_{\alpha/2}$ is the $1-\alpha/2$ quantile from the $t$-distribution (the degrees of freedom depend on your assumption about equal variance).

Before jumping to an example, take a moment to look at the above null and alternative hypotheses. 
The null hypothesis states $\mu_A=\mu_B$, which effectively says there is no difference in the mean between the two groups. 
That is, whether the data came from group A or group B it does not provide any explanation about the variability in the data.
However, the alternative says that there is a difference and that knowing the group from which the data came helps explain some of the observed variability. 

Now for an example, consider comparing the ACT scores for incoming freshmen in 1996 to 2000 using the university admission data. Perhaps we have a theory that ACT scores were higher in 1996 than 2000. 
That is, we formally wish to test $$H_0:\ \mu_{1996} = \mu_{2000} ~~~\textrm{versus}~~~ \mu_{1996} > \mu_{2000}$$

First we need to select a subset of the data corresponding to those two years.

```{r ch1-27}
uadata.trim <- uadata |>
  filter(year %in% c(1996, 2000) )
```

Here we `filter()` the `uadata` so that only years in the vector `c(1996, 2000)` are included. Now that we have a proper subset of the data, we can compare the ACT scores between the two years and build a 98% confidence interval for the mean difference.

```{r ch1-28}
t.test(act ~ year, data=uadata.trim, conf.level=0.98, 
       alternative="greater", var.equal=TRUE)
```

We are 98% confident that the mean difference between 1996 ACT scores and 2000 ACT scores is in the interval (-0.792, $\infty$).
Further, we lack significant evidence ($p$-value $\approx 0.3817$) to conclude that the average ACT score in 1996 is greater than that in 2000 for incoming freshmen (also note that the value 0 is included in the confidence interval).

#### Coding Notes

A few things about the code above.

* We use the notation `act ~ year`, this tells R that the `act` scores are a function of `year`. This is known as a `formula` and will be important throughout the course.
* We specify the data set to be used to perform the test (`data=uadata.trim`).
* We specify the confidence level (by default it is 0.95) with `conf.level=0.98` and the alternative hypothesis (by default it is a not-equal test) with `alternative=greater`.
* Lastly, we set `var.equal=TRUE` so the standard error (and degrees of freedom) are calculated under the assumption that the variance of the two populations is equal (pooled variance).

