

```{r ch4-setup, include=FALSE}
library(knitr)
library(kableExtra)
library(tidyverse)
library(broom)
library(emmeans)
library(patchwork)
library(ggfortify)
options(dplyr.summarise.inform = FALSE,
        warn = -1)
```


# Advanced Designs

The previous chapters have established the groundwork of experimental designed. 
We discussed the key elements of experimental units, factors, response variables, treatments, confounding influences and the purpose of replication and randomization. 
This former work just scratches the surface of the field of experimental design and the analyses associated with those designs. 
The objective of this chapter includes:

* Understanding the basic structure of factorial designs.
* Mixing blocking and factorial components in a design.
* Understanding the basic structure of within-subject designs.
* Introducing *random effects* into a model.
* **Building an appreciation** for the complexity of experimental design.

## Higher order factor models

In the previous two chapters we have seen a One-Way and Two-Way ANOVA.
In the former, a single factor consisted of $k$-levels (and thus $k$-treatments) while in the latter the first factor may consist of $k$ levels and the second factor $j$ levels (and therefore $k\times j$ treatments).
This idea can, of course, be expanded to even more terms, consider the following example.

### Example of Three-factor design

**Example.** Marketing research contractors (@KutnerText). 

A marketing research consultant evaluated the effects of free schedule, scope of work and type of supervisory control on the quality of work performed under contracts by independent marketing research agencies.
The design of this experiment can be summarized as follows

$$\begin{array}{ll} 
\textbf{Factor} & \textbf{Factor Levels} \\
\hline
\textrm{Fee level} & \textrm{High} \\
 & \textrm{Average} \\
 & \textrm{Low}\\
 \hline
\textrm{Scope} & \textrm{In house} \\
 & \textrm{Subcontracted} \\
\hline
\textrm{Supervision} & \textrm{Local Supervisors}\\
 & \textrm{Traveling Supervisors} \\
\end{array}$$

The quality of work performed was measured by an index taking into account several characteristics of quality. 

There are **three factors** in this experiment, with 3, 2 and 2 levels for each factor.
So the number of treatments is $3\times 2\times 2 = 12$ and can be summarized in the following table.

```{r three_way_table, echo=FALSE}
tab <- data.frame("Fee Level" = c("High", "Average", "Low"),
                  "In house" = c("Treatment 1", "Treatment 5", "Treatment 9"),
                  "Subcontracted" = c("Treatment 2", "Treatment 6", "Treatment 10"),
                  "In house" = c("Treatment 3", "Treatment 7", "Treatment 11"),
                  "Subcontracted" = c("Treatment 4", "Treatment 8", "Treatment 12"))
names(tab) <- c("Fee Level", "In house", "Subcontracted", "In house", "Subcontracted")
kable(tab) |> kable_styling(full_width=FALSE) |>
  add_header_above(c(" " = 1, "Local Supervisers" = 2, "Traveling Supervisors" = 2)) |>
  add_header_above(c(" " = 1, "Supervision" = 4))
```

Recall the main idea from two-factor experiments--the interaction terms are typically very important. 
In this three factor design we have the three main effects (Fee level, Scope and Supervision) but also have pairwise interactions between these variables and an interaction among all three.
From a statistical modeling perspective, we are now essentially fitting the model 

$$Y_{ijkm} = \mu + \alpha_i + \beta_j + \gamma_k + (\alpha\beta)_{ij} + (\alpha\gamma)_{ik} + (\beta\gamma)_{jk} + (\alpha\beta\gamma)_{ijk} + \varepsilon_{ijkm}$$

where

* $Y_{ijkm}$ is the $m^\mathrm{th}$ observation in the $i^\mathrm{th}$ level of the first factor, the $j^\mathrm{th}$ level of the second factor and the $k^\mathrm{th}$ level of the third factor.
* $\mu$ is an overall mean.
* Fee level main effect $\alpha_i$.
* Scope main effect $\beta_j$.
* Supervision main effect $\gamma_k$.
* Fee level and Scope interaction effect $(\alpha\beta)_{ij}$.
* Fee level and Supervision interaction effect $(\alpha\gamma)_{ik}$.
* Scope and Supervision interaction effect $(\beta\gamma)_{jk}$.
* Fee level, Scope and Supervision interaction effect $(\alpha\beta\gamma)_{ijk}$.
* $\varepsilon_{ijkm}$ is the random error term.

As you can see this model has grown greatly compared to to the two-way ANOVA examples from before. 

**What will happen if there are four factors?**

In general, if an experiment consisted of $m$ different factors, we would potentially have $2^m - 1$ terms in our model.
Thus with four factors you are looking at $2^4 - 1 = 16-1=15$ terms in your ANOVA decomposition.
With five factors, 31 terms. 
This will quickly grow out of control.
We may run out of Greek letters before long!

Fortunately, for higher order factor problems there are statisticians available to help formulate a better design and potentially *optimize* some of the variables of study. 
We leave out the details of these topics in this text but want to provide insight that these advanced designs and analyses can be done. 
Please consult a statistician before your experiment if you are planning a complex design!!

### Three-factor ANOVA Example

**Example continued.** Below we complete the example involving the marketing research contractors with the data available in the `marketingResearch.txt` file on the book data repository. 

First we read in the data and make sure R properly treats the factor variables.
From (@KutnerText) we have three fee levels corresponding as 1=High, 2=Average, 3=Low.
The two scope levels are 1=all work in-house, 2=subcontracted, and the two supervision levels are 1=local supervisors, 2=traveling supervisors. 

```{r ch4marketingData-noeval, eval=FALSE}
data_site <- "https://tjfisher19.github.io/introStatModeling/data/marketingResearch.txt"
marketing <- read_table(data_site)
glimpse(marketing)
```

```{r ch4marketingData-noEcho, echo=FALSE}
marketing <- read_table("docs/data/marketingResearch.txt", col_types=cols() )
glimpse(marketing)
```

```{r ch4marketingData}
marketing <- marketing |>
  mutate(Fee = factor(Fee, levels=1:3,
                      labels=c("High", "Average", "Low")),
         Scope = factor(Scope, levels = 1:2,
                        labels=c("In-house", "Subcontracted")),
         Supervision = factor(Supervision, levels=1:2,
                              labels=c("Local", "Traveling") ) )
```

#### EDA for three-factors

We have replication within each treatment (4 replicates for each of the $3\times 2\times 2 = 12$ unique treatments for $n=48$ total observations).
Any EDA needs to account for the 12 unique treatments over the 3 different factor variables.
Below we make a visual summarizing the mean response for each unique treatment, linking all profiles together for both supervision and scope -- these two variables were chosen as each factor has two unique values and this limits the number of colors and linetypes needed in the plot.

```{r ch4marketingEDA, fig.align='center', fig.cap='Interaction plot for the 3-factor experiment looking at consulting quality as a function of fee, scope and supervision.', out.width='60%', fig.asp=0.55}
ggplot(marketing, aes(x=Fee, y=Quality, color=Scope, linetype=Supervision,
                      group=interaction(Scope, Supervision)) ) + 
    stat_summary(fun=mean, geom="point") + 
  stat_summary(fun=mean, geom="line") +
  theme_minimal()
```
Visually, it appears the High and Average fee types have the highest quality regardless of scope & supervision. 
We also see that the Fee type does not appear to interact with the Scope and Supervision variables as all four lines look reasonably parallel.
The connection between Scope and Supervision is more complex; note the alternating colors when looking at the lines while the linetypes do not alternate in the same way. 
This may be an indication of some sort of interaction between these two variables.

#### Fitting a three-factor model

We fit the ANOVA model.
Here we used the notation `Fee*Scope*Supervision` so R will fit a full factorial model (*that is*, all main effects and interaction terms). 

```{r ch4marketingANOVA, fig.align='center', fig.cap='Residual plots for the Three-Way ANOVA studying the twelve unique treatments on the quality of consultants - Top-left a Residuals vs Fitted Plot; Top-right a Normal Q-Q plot of the residuals; Bottom-left the Scale-Location plot; and Bottom-right a Residuals vs Leverage plot.', fig.asp=0.9, out.width='70%'}
marketing.anova <- aov(Quality ~ Fee*Scope*Supervision, data=marketing)
autoplot(marketing.anova)
```

The residual plot in \@ref(fig:ch4marketingANOVA) shows that the constant variance assumption is generally met (no concerning patterns in either the Residuals vs Fitted or Scale-Location plot). 
The Normal Q-Q plot does show some deviation from the 45-degree line but here we must make note of how it deviates.

* We do see deviation in both tails
* The *observed* or *sample* residuals generally have smaller $Z$-scores than is expected, this indicates the data is *light tailed* (less data in the extremes of the distribution).
* We also note that the deviations are symmetric (on both sides)

Given we have no skewness and the data is light tailed, the central limit theorem can justify the use of ANOVA in this setting.
That is, even if the residuals are not exactly normally distributed, the sample size is adequate to provide some justification to the result.

Now look at the ANOVA output.

```{r ch4markingSummary}
summary(marketing.anova)
```

Similar to the two-factor analysis, **always look at the most complex interaction term first.**
Here the most complex term is the `Fee:Scope:Superversion` term. 
That three factor interaction term is not significant ($F$-statistic of 0.267 on 2 and 36 degrees of freedom, $p$-value=0.767), thus we can proceed to the pairwise interactions.

* The interaction between Scope and Supervision is significant ($F$-stat of 77.749 on 1 and 36 degrees of freedom, $p$-value$\approx 10^{-10}$).
* Both interaction terms involving the Fee level factor are not significant
   + The `Fee:Scope` term is not significant ($F$-stat of 0.108 on 2 and 36 degrees of freedom, $p$-value=0.898)
   + The `Fee:Supervision` term is not significant ($F$-stat of 0.053 on 2 and 36 degrees of freedom, $p$-value=0.948), 
   
Thus, since the interaction of the Fee variable with any other factor variable does not appear to significantly influence the response, we can analyze its main effect variable.
The Fee main effect influences the quality ($F$-stat of 679.336 on 2 and 36 degrees of freedom, $p$-value$\approx10^{-16}$).

Overall, we can conclude that some combination of Scope and Supervision influence quality. 
We also conclude that the different Fee types appear to influence Quality.

#### Multiple comparisons

In general, multiple comparisons analysis can get very complex when three or more factors are involved.
In this case, we get lucky

* Fee` is significant, but *only* as a main effect.
* The `Scope` \& `Supervision` interaction term is significant, but it is only 2 variables and we covered that analysis in section \@ref(twoWayInteraction).

First we consider the effects of Fee on the quality index

```{r, fig.align='center', fig.cap='Tukey HSD comparison of three levels of the Fee variable on consultant quality. The Low fee option results in a lower quality than High or Average, which statistically are equivalent.', out.width='60%', fig.asp=0.55}
marketing.mc.fee <- emmeans(marketing.anova, ~Fee)
plot(contrast(marketing.mc.fee, "pairwise"))
```
We are building intervals of comparisons, so we look for the inclusion or exclusion of the value zero. We see that the "Low" fee level is statistically different than both "High" and "Average", while the later two are not statistically different.

To analyze the Scope & Supervision influence on Quality we must do so conditionally. 

```{r, fig.align='center', fig.cap='Tukey HSD comparison of two levels of the Scope variable on consultant quality conditioned on Supervision. In both settings of supervision, the In-house scope results in a higher quality than subcontracted. It also appears this difference is greater when Traveling Supervision is used.', out.width='60%', fig.asp=0.65}
marketing.mc.scope <- emmeans(marketing.anova, ~Scope|Supervision)
plot(contrast(marketing.mc.scope, "pairwise"))
```

### Summarizing higher order factors


## Combling Blocks and Factorial Models


```{r}
dentalPain <- read_csv("docs/data/dentalPain.csv")
glimpse(dentalPain)
```

```{r}
dentalPain <- dentalPain |>
  mutate(PainLevel = factor(PainLevel),
         Codeine = factor(Codeine, levels=1:2,
                          labels=c("Sugar Pill", "Codeine capsule")),
         Acupuncture = factor(Acupuncture, levels=1:2,
                              labels=c("Inactive Acupuncture",
                                       "Active Acupuncture")))
```

```{r}
dentalpain.anova <- aov(Relief ~ PainLevel + Codeine*Acupuncture,
                        data=dentalPain)
```

```{r}
autoplot(dentalpain.anova)
```
```{r}
summary(dentalpain.anova)
```

```{r}
dentalPain <- dentalPain |>
  mutate(Treatment = paste(Codeine, Acupuncture))
```

```{r}
dentalPain.anova2 <- aov(Relief ~ PainLevel + Treatment,
                         data=dentalPain)
```

```{r}
summary(dentalPain.anova2)
```

```{r}
sort(unique(dentalPain$Treatment))
```

```{r}
dental.em <- emmeans(dentalPain.anova2, ~Treatment)
contrast(dental.em, "dunnett", ref=4)
```




## Within-subject designs

In the previous two chapters, you have seen different experimental designs where a factor(s) levels are manipulated by the researcher in order to measure a response. 
In simpler design settings, the factors that are manipulated are of the "between-subjects" type.
The one-way ANOVA example in chapter 2 on drug effects on alertness is illustrative of this: each participant (experimental unit) had exactly one drug randomly assigned to them.
As such, there is only one observation per experimental unit in the design. 
Drug was a **between-subjects factor**.

*Why is that important to note?*
Because, as you may realize by now, that assures that the observations in the data are all independent.
It will be elaborated upon later in this course how crucial this assumption is to the validity of any statistical inferences we draw from the tests/CIs we derive.
In fact, independence is more important to inference validity than either normality or constant variance assumptions.

Now, contrast this with the **paired $t$-test** (see section 2.4).
An example was discussed that involved giving a sample of 20 people a drug treatment in order to study its effect on the Cholesterol of the patients. 
Think of that problem from a design of experiments point of view: what was/were the factor(s) that were manipulated by the researcher?

* It is not the drug treatment! Why? Because in the experiment, only one drug was studied: there was no "comparison" drug or placebo used to form a basis for comparison. As such, the drug treatment in that problem was not a variable at all: it was a constant.
* So what is left? The answer is TIME. The researcher measured Cholesterol at two time points on each patient (experimental unit): [1] just before the start of the drug treatment, and [2] 45 days later. The factor here is time, and it has two levels. Time is a **within-subject factor**.

In a designed experiment that involves several measurements made on the same experimental unit over time, the design is called a **repeated measures design**.

The important thing to recognize here is that each patient provided two observations, not one. 
It stands to reason that *what a particular subject’s Cholesterol is after 45 days will be related to their starting Cholesterol before the drug treatment*. 
That is the layman’s way of saying that the measurements within a subject are **not independent** (*i.e.*, a patient is not independent of themselves -- the observations are correlated).
If in any analysis we ignored this basic truth, that analysis would be seriously flawed.

So how did we handle this before?
A simple strategy which is common with paired data comparisons is to calculate each subject’s difference score (yielding a single difference measurement per subject) and then performing a one-sample $t$-test on the differences.
Each difference score is independent because each came from a different randomly chosen subject.
This is what is done in the paired $t$-test analysis in section 2.4. 

### Blocks revisited: an approach to handling within-subjects factors

The notion and purpose of blocking in an experimental design was introduced in section 3.1.
Recall the definition of block terms as a relatively homogeneously set of experimental material.
We saw two examples where subjects serve as blocks. 
What this means for us now is that using a block design approach in analysis is 'a way' to handle correlated (non-independent) observations such as in the paired $t$-test problem above. 
We illustrate that here.

**Example.** Carbon Emissions (originally in @WeissText).

The makers of the MAGNETIZER Engine Energizer System (EES) claim that it improves the gas mileage and reduces emissions in automobiles by using magnetic free energy to increase the amount of oxygen in the fuel for greater combustion efficiency.
Test results for 14 vehicles that underwent testing is available in the file `carCOemissions.csv`.
The data includes the emitted carbon monoxide (CO) levels, in parts per million, of each of the 14 vehicles tested before and after installation of the EES.

This is a classic before-after (Paired design) study. Below is some code to implement the test.

First, we quickly recap the Paired $t$-test result from Section REF

```{r, eval=FALSE}
www <- "https://tjfisher19.github.io/introStatModeling/data/carCOemissions.csv"
vehicleCO <- read_csv(www, col_types = cols())
vehicleCO
```

```{r ch4-coData, echo=FALSE}
vehicleCO <- read_csv("docs/data/carCOemissions.csv", col_types = cols())
vehicleCO
```

and the paired $t$-test

```{r ch4-coData-ttest}
t.test(vehicleCO$before, vehicleCO$after, paired=TRUE)
```

Now, we will perform this same analysis using a Block ANOVA. 
The data is structure that 2 observations (the before and after) are on each row of the data frame.
We will reshape this data into a tall, or long, format.

```{r ch4-coData-reshape}
vehicleCO_tall <- vehicleCO |>
  pivot_longer(cols=c(before, after),
               names_to="Time",
               values_to="Emitted_CO")
head(vehicleCO_tall)
```

Note the difference in the layout of the data here as compared to the original `vehicleCO` used in the paired $t$-test: each observation occupies a single row of the data file now.
There is an identifier variable (`id`) that identifies which subject every observation belongs to. 
The variable Time `refers` to the time factor and `Emitted_CO` records the emitted Carbon Monoxide levels (ppm).

We will analyze this same data (addressing the same research question) by using a **block design approach**:

```{r ch4-2}
vehicleco.block.aov <- aov(Emitted_CO ~ factor(id) + Time, data=vehicleCO_tall)
summary(vehicleco.block.aov)
```

Look at the $F$-test for the time factor (labeled `Time` here).
This result is identical to the result of the paired t-test above ($F$ = 9.897, $\textrm{df}_1$ = 1, $\textrm{df}_2$ = 13, $p$-value $= 0.00773$).
Note, $\sqrt{9.897} =$ `r round(sqrt(9.897), 3)`, the value of the paired $t$-test statistic.

*How does the block approach work here?* You will note that the test of Time (the within subjects effect) uses the residual effect as the error term. 
Because the blocks are accounted for in the model (and here, the blocks constitutes the overall between-vehicle variability source), the only variation left over for the "residuals" is the variability due to the within-vehicle differences.
In essence, there are **two components** to the random error now:

* Aggregated variability between vehicles (block effect)
* Remaining unexplained (random) variability after accounting for aggregate vehicle-to-vehicle differences and aggregated time-to-time differences. We will call this leftover stuff "pure" error.

For a correct analysis, it is important that each factor of interest in the study be tested against the proper component of the error. Here, **we test a within-subjects effect against the random error component dealing with "pure" random error**.

Perhaps **a more general approach** to within-subjects analysis is to specify to R the fact that there are layers of random variability in the data introduced by the fact that we have more than one measurement per experimental unit. 
This can be done using an `aov` model specification that explicitly cites these layers with an `Error` option. 
The model specification in R is generally

    `aov(response ~ between-subjects factor structure + Error(EU variable/WS factor)`

In the present example, this would be as follows:

```{r ch4-3}
WS_aov <- aov(Emitted_CO ~ Time + Error(factor(id)/Time), data=vehicleCO_tall)
summary(WS_aov)
```

Again we see the same result. It turns out that for more complex studies this general approach is preferred.



### A more involved repeated measures case study

If we can handle within-subjects factors using blocks, why not just stick with that all the time?  
Well, the reason is that repeated measures designs can get complicated: multiple time measurement points (more than just two), the addition of between-subjects factors, etc...
The following case study example illustrates this.

**Example.** Cholesterol study (from the University of Sheffield [@sheffield2021]).

A study tested whether cholesterol was reduced after using a certain brand of margarine as part of a low fat, low cholesterol diet. The subjects consumed on average 2.31g of the active ingredient, *stanol ester*, a day. This data set `cholesterolreduction.csv` contains information on 18 people using margarine to reduce cholesterol over three time points.  


```{r ch4tab, echo=FALSE}
tab <- data.frame("Variable Name"=c("ID", "Before", "After4weeks", "After8weeks", "Margarine"),
                  `Variable` = c("Participant Number", "Cholesterol before the diet (mmol/L)", "Cholesterol after 4 weeks on diet (mmol/L)", "Cholesterol after 8 weeks on diet (mmol/L)", "Margarine type A or B"),
                  `Data type` = c("", "Scale", "Scale", "Scale", "Binary") )
names(tab) <- c("Variable Name", "Variable", "Data type")
kable(tab)
```


```{r, eval=FALSE}
"carCOemissions.csv"
chol <- read_csv("https://tjfisher19.github.io/introStatModeling/data/cholesterolreduction.csv", col_types=cols())
glimpse(chol)
```

```{r ch4-4, echo=FALSE}
chol <- read_csv("docs/data/cholesterolreduction.csv", col_types=cols())
glimpse(chol)
```

There are two factors of interest here: 

* Time points (3 levels within-subjects): before, after4weeks, after8weeks)
* Margarine type (2 level between-subjects): A or B

Note that the data is currently in wide form.
We need to convert it to tall form:

```{r ch4-5}
chol_tall <- chol |> 
  pivot_longer(cols=Before:After8weeks, 
               names_to="time.pt", 
               values_to="chol.level")
head(chol_tall)

chol_tall <- chol_tall |>
  mutate(time.pt = factor(time.pt, 
                          levels=c("Before", "After4weeks", "After8weeks"),
                          labels=c("Before", "After 4 weeks", "After 8 weeks")))
```

Note here we have changed the variable `time.pt` to be a factor and we order the factors levels so they are contextually sequential (rather than alphabetical order).


#### Exploratory Data Analysis

We are now ready for some EDA. A table of descriptive statistics by combinations of time and margarine group is our first cut:

```{r ch4-6}
chol.summary <- chol_tall |>
  group_by(Margarine, time.pt) |>
  summarise(Mean = mean(chol.level),
            SD = sd(chol.level),
            SE = sd(chol.level)/sqrt(length(chol.level)),
            n = length(chol.level))
kable(chol.summary)
```

We generally observe that cholesterol levels with margarine A appear lower than with margarine B. 
Also, there does appear to be some reduction in cholesterol level using both products: however, there also appears to be more unexplained variability ("noise") in the cholesterol measurements under margarine A. 
We can further check this using some data visualizations. 
First let us try boxplots (not appropriate given the design) as seen in Figure \@ref(fig:ch4-7).

```{r ch4-7, fig.align='center', fig.cap='Boxplot distribution of the Cholesterol level as a function of Treatment and Time of measurement.',out.width='80%', fig.asp=0.55}
ggplot(chol_tall) +
  geom_boxplot(aes(x=time.pt, y=chol.level)) +
  facet_wrap(~Margarine) +
  xlab("Time Point of Measurement") +
  ylab("Cholesterol level (mmol/L)")
```

We see in Figure \@ref(fig:ch4-7) that perhaps cholesterol levels are generally more consistent under margarine B, and that it is just some outlier(s) that are the issue. 
However, this plot does not reveal a crucial piece of information: **the responses are correlated over time**. 
A better depiction would allow us to see how each individual subject’s cholesterol changed over the span of the experiment as seen in Figure \@ref(fig:ch4-8).

```{r ch4-8, fig.align='center', fig.cap='Profile plots of the Cholesterol level as a function of Treatment and Time of measurement.',out.width='80%', fig.asp=0.55}
# Plot the individual-level cholesterol profiles
ggplot(chol_tall, aes(x=time.pt, y=chol.level, colour=Margarine, group=ID)) +
  geom_line() + 
  geom_point(shape=21, fill="white") + 
  facet_wrap(~Margarine) +
  xlab("Time Point of Measurement") +
  ylab("Cholesterol level (mmol/L)") +
  ggtitle("Subject-level cholesterol profiles by Margarine group")
```

In Figure \@ref(fig:ch4-8) we see some things more clearly. 
The general pattern over time appears to be pretty consistent for every experimental unit (visually, each "line" in the plot). 
There are two individuals in margarine B whose cholesterol level seems unusually higher than the others: this might warrant further investigation. 
Now we provide an summary profile of the two treatments in \@ref(fig:ch4-9).

```{r ch4-9, fig.align='center', fig.cap='Average Profile of Cholesterol level by Treatment in Time.', out.width='70%', fig.asp=0.55}
# we include a position_dodge elements to avoid overlap
ggplot(chol.summary, aes(x=time.pt, y=Mean, colour=Margarine)) + 
  geom_errorbar(aes(ymin=Mean-SE, ymax=Mean+SE), width=.1, position=position_dodge(0.3)) +
  geom_line(aes(group=Margarine), position=position_dodge(0.3)) +
  geom_point(position=position_dodge(0.3)) +
  xlab("Time Point of Measurement") +
  ylab("Cholesterol level (mmol/L)") +
  ggtitle("Mean Cholesterol level (with Standard Error bars)")
```



In the summary version (\@ref(fig:ch4-9)) we see that margarine A Cholesterol levels appear lower than margarine B. 
It is worth noting that the group assigned to margarine level A had lower cholesterol "before" the study began. 

#### Preliminary inferential analysis

Of more concern to a fully correct analysis is addressing why there appears to be a difference between the margarine groups at the "Before" stage.  At that point, no margarines had been administered, so if the subjects had been truly randomly assigned to the margarine groups (as they should be in a properly designed experiment!), we would have expected consistent cholesterol readings between the groups at that point.  This can be formally tested as follows, with a simple independent samples $t$-test comparing the mean cholesterol levels of the A and B groups at the "Before" stage:

```{r ch4-10}
BeforeDataOnly <- chol_tall |>
  filter(time.pt == "Before")
t.test(chol.level ~ Margarine, data=BeforeDataOnly, var.equal=TRUE)
```

The result is not statistically significant ($p$-value = 0.1932).  That is actually good news here! Although visually it appears different it is not statistically different. 

#### A conditional repeated measures analysis

To get started, lets do a within-subjects (repeated measures) ANOVA on the Margarine A data only.  The goal here would be to see if there is an effective reduction in mean cholesterol over the eight-week trial by using margarine A.  Here goes: follow the code!

```{r ch4-11}
MargarineA <- chol_tall |> 
  filter(Margarine == "A")
RMaov_MargA <- aov(chol.level ~ time.pt + Error(factor(ID)/time.pt), data=MargarineA)
summary(RMaov_MargA)
```

The test of interest is labeled time.pt.  It is significant ($F$ = 106.3, $\textrm{df}_1$ = 2, $\textrm{df}_2$ = 16, $p$-value = $5.7 \times 10^{-10}$).  What this means is that, using Margarine A, there is significant evidence to conclude that the *true mean cholesterol level differs between at least two of the time points*.  Of course, that result does not mean that the margarine is effective at reducing cholesterol levels.  All it means is that the mean cholesterol level is changing at some point over the experiment.  Since there are three time points, a multiple comparison procedure is warranted with the output below and in Figure \@ref(fig:ch4-12).

```{r ch4-12, fig.align='center', fig.cap='Comparing the effects of time within the Margarine A treatment groupu. We see there is no statistical difference between 4 and 8 weeks after treatment begins.', out.width='70%', fig.asp=0.55}
MargA.mc <- emmeans(RMaov_MargA, ~ time.pt)
contrast(MargA.mc, "pairwise")
plot(contrast(MargA.mc, "pairwise"))
```

The significant reductions in mean cholesterol level under Margarine A occurs right after the start of the experiment.  The mean cholesterol level is reduced by 0.485 (SE = 0.041) by week 4 ($p$-value < 0.0001); and by 0.546 (SE = 0.041) by week 8 ($p$-value < 0.0001).  There is no appreciable change in mean cholesterol level between weeks 4 and 8 ($p$-value = 0.3229).

#### Full analysis: Formal Comparison of Margarines

It is always critically important to know the structure of the data, which here stems from the design of the experiment.  In this case, we cannot just "compared the margarines" because there is a subject-level time profile under each margarine.  Since the same time points are observed under each margarine group, the factors margarine and `time.pt` are crossed effects, which means we have a *factorial data structure* (see section 3.2).  The "wrinkle" here is that one of the two factors is a between-subjects factor (Margarine) and the other is within-subjects (`time.pt`).  But, we can handle this in `aov`: 

```{r ch4-13}
RMaov_all <- aov(chol.level ~ Margarine + time.pt + Margarine:time.pt + 
                              Error(factor(ID)/time.pt), data=chol_tall)
summary(RMaov_all)
```

As with the factorials models before, the first thing we must check is the significance of the interaction term.  If this is significant, it means that the effect of time on the mean cholesterol is different between the two margarines.  That is the case here ($F$ = 4.777, $\textrm{df}_1$ = 2, $\textrm{df}_2$ = 32, $p$-value = 0.0153).  As a result, we should do our multiple comparisons **by conditioning on one of the two factors**.  Experimental context will dictate which of the two factors is the more reasonable to condition on.  Here, I choose to condition on Margarine group, so I will get a time-based comparison of mean cholesterol level, but specific to each Margarine group:

```{r ch4-14, fig.align='center', fig.cap='Graphical exploration of all treatments noting that there is no statistical difference between 4 and 8 weeks after the study has began regardless of Margarine treatment.', out.width='70%', fig.asp=0.75}
RMall.mc <- emmeans(RMaov_all, ~ time.pt | Margarine)
contrast(RMall.mc, "pairwise")
plot(contrast(RMall.mc, "pairwise"))
```

Even though at first glance the time-based comparisons look about the same between the two margarines, the point estimates and plots reveal the story: the reductions in mean cholesterol following the initiation of the study are larger for margarine B (mean=0.7111 mmol/L, SE=0.043 by week 8) than for margarine A (mean=0.546 mmol/L, SE=0.043 by week 8).  

### Further study

More thorough analysis of within-subjects and repeated measures designs are possible using advanced methods beyond the scope of this course.  These methods include:

* processes that allow the analyst to estimate the degree of correlation between repeated measurements
* processes that allow one to model different correlation structures based upon experimental realities in the data collection process

For example, suppose we had a repeated measures experiment with subjects measured across four time points: Week 1, 2, 3 and 8.  Because of the time spacing, we might expect that measurements between adjacent weeks to be more similar within a subject than weeks far removed in time. For example, it is reasonable to expect that results in week 2 are more strongly correlated to week 1 results than would be the correlation between weeks 3 and 1.  Week 8’s results might be so far removed from the first few weeks that the results there might be safely presumed to be nearly independent of earlier results.  We  could even be more restrictive, using an approach that assumes only adjacent time point results are correlated, but beyond that, assume independence.

Subtleties like this can be modeled and checked using advanced modeling techniques.

## Random Effect Models
